$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-09-08 paper/paper.tex
Move subject congruence to the body

\subparagraph*{Framing}
asserts that the well-typedness of a process is independent of its leftover resources.

@@

\subparagraph*{Weakening}
states that inserting a new variable into the context preserves the well-typedness of a process as long as the usage annotation of the inserted variable is preserved as a leftover.
Let $\func{ins}_i$ insert an element into a context at position $i$ --- for simplicity, we use it both to insert types into typing contexts and usage annotations into usage contexts.
\begin{nitheorem}[Weakening]
  \label{thm:weakening}
  Let $P$ be well typed in $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Then, lifting every variable greater than or equal to $i$ in $P$ is well typed in
  $\types{\func{ins}_i \; t \; \gamma}{\func{ins}_i \; x \; \Gamma}{\func{lift}_i \; P}{\func{ins}_i \; x \; \Xi}$.
\end{nitheorem}

\subparagraph*{Strengthening}
states that removing an unused variable preserves the well-typedness of a process.
Let $\func{del}_i$ delete the element at position $i$ from a context --- for simplicity, we use it both to delete types from typing contexts and usage annotations from usage contexts.
\begin{nitheorem}[Strengthening]
  \label{thm:strengthening} 
  Let $P$ be well typed in $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Let $i$ be a variable not in $P$, such that $\Unused_i \; P$.
  Then lowering every variable greater than $i$ in $P$ is well typed in $\types{\func{del}_i \; \gamma}{\func{del}_i \; \Gamma}{\func{lower}_i \; P}{\func{del}_i \; \Xi}$.
\end{nitheorem}

@@

\subparagraph*{Exchange}
of two variables preserves the well-typedness of a process.
We extend $\func{exchange}_i$  introduced in \autoref{structural-congruence} to exchange types in typing contexts and usage annotations in usage contexts.
\begin{nitheorem}[Exchange]
  \label{thm:exchange}
  Let $P$ be well typed in $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Then, $\types{\func{exchange}_i \; \gamma}{\func{exchange}_i \; \Gamma}{\allowbreak \func{exchange}_i \; P}{\func{exchange}_i \; \Xi}$.
\end{nitheorem}

@@

\subsection{Towards Subject Reduction}
\subparagraph*{Subject Congruence}
states that applying structural congruence rules to a well typed process preserves its well-typedness.

@@

\begin{proof}[Proof (Sketch)]
  The proof is by induction on \textsc{Equals} $\eq{}$.
  For the nontrivial base cases and their symmetric variants, we apply a combination of \autoref{thm:framing}, \autoref{thm:strengthening}, \autoref{thm:weakening} and \autoref{thm:exchange}.
  Further details can be found in \autoref{app:struct} and the full proof in%
\begin{anonsuppress}
\cite{Zalakain2020Agda}
\end{anonsuppress}
.
\end{proof}

\subparagraph*{Substitution}
\autoref{thm:substitution-generalization} proves a generalised version of substitution, where the substitition $\subst{P}{j}{i}$ is on any variable $i$.
\autoref{thm:substitution} instantiates the generalised version to the concrete case where $i$ is the most recently introduced variable $\constr{0}$, as required by subject reduction.

@@

\subparagraph*{Subject Reduction}
states that if $P$ is well typed and it reduces to $Q$, then $Q$ is well typed.
In the \picalc{} we distinguish between a reduction $P \reduce{\constr{internal}} Q$ on a channel internal to $P$, and a reduction $P \reduce{\constr{external} \; i} Q$ on a channel $i$ external to $P$ (refer to \autoref{operational-semantics}).

>>>>

\paragraph*{Framing}
asserts that the well-typedness of a process is independent of its leftover resources.

@@

\paragraph*{Weakening}
states that inserting a new variable into the context preserves the well-typedness of a process as long as the usage annotation of the inserted variable is preserved as a leftover.
Let $\func{ins}_i$ insert an element into a context at position $i$ --- for simplicity, we use it both to insert types into typing contexts and usage annotations into usage contexts.
\begin{nitheorem}[Weakening]
  \label{thm:weakening}
  Let $P$ be well typed in $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Then, lifting every variable greater than or equal to $i$ in $P$ is well typed in
  $\types{\func{ins}_i \; t \; \gamma}{\func{ins}_i \; x \; \Gamma}{\func{lift}_i \; P}{\func{ins}_i \; x \; \Xi}$.
\end{nitheorem}

\paragraph*{Strengthening}
states that removing an unused variable preserves the well-typedness of a process.
Let $\func{del}_i$ delete the element at position $i$ from a context --- for simplicity, we use it both to delete types from typing contexts and usage annotations from usage contexts.
\begin{nitheorem}[Strengthening]
  \label{thm:strengthening} 
  Let $P$ be well typed in $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Let $i$ be a variable not in $P$, such that $\Unused_i \; P$.
  Then lowering every variable greater than $i$ in $P$ is well typed in $\types{\func{del}_i \; \gamma}{\func{del}_i \; \Gamma}{\func{lower}_i \; P}{\func{del}_i \; \Xi}$.
\end{nitheorem}

@@

\paragraph*{Exchange}
of two variables preserves the well-typedness of a process.
We extend $\func{exchange}_i$  introduced in \autoref{structural-congruence} to exchange types in typing contexts and usage annotations in usage contexts.
\begin{nitheorem}[Exchange]
  \label{thm:exchange}
  Let $P$ be well typed in $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Then, $\types{\func{exchange}_i \; \gamma}{\func{exchange}_i \; \Gamma}{\allowbreak \func{exchange}_i \; P}{\func{exchange}_i \; \Xi}$.
\end{nitheorem}

@@

\subsection{Towards Subject Reduction}
\paragraph*{Subject Congruence}
states that applying structural congruence rules to a well typed process preserves its well-typedness.

@@

\begin{proof}[Proof (Sketch)]
  The proof is by induction on \textsc{Equals} $\eq{}$.
  Here we only consider those cases that are not purely inductive: the base cases for $\constr{struct}$ and their symmetric variants.
\begin{anonsuppress}
Full proof in \cite{Zalakain2020Agda}.
\end{anonsuppress}
  We proceed by induction on \textsc{StructCong} $\eqeq$:
  \begin{itemize}
    \item
      Case $\constr{comp-assoc}$: trivial, as leftover typing is naturally associative.
    \item
      Case $\constr{comp-sym}$ for $\comp{P}{Q}$: we use framing (\autoref{thm:framing}) to shift the output context of $P$ to the one of $Q$; and the input context of $Q$ to the one of $P$.
    \item
      Case $\constr{comp-end}$: trivial, as the typing rule for $\PO$ has the same input and output contexts.
    \item
      Case $\constr{scope-end}$: we show that the usage annotation of the newly created channel must be $\lz$, making the proof trivial.
      In the opposite direction, we instantiate the newly created channel to a type $\unit$ and a usage annotation $\lz$.
    \item
      Case $\constr{scope-ext}$ for $\new \comp{P}{Q}$: we to show that $P$ preserves the usage annotations of the unused variable (\autoref{lm:types-unused}) and then use strengthening (\autoref{thm:strengthening}).
      In the reverse direction, we use weakening (\autoref{thm:weakening}) on $P$ and show that lowering and then lifting $P$ results in $P$ (\autoref{lm:lower-lift}).
    \item
      Case $\constr{scope-comm}$:
      we use exchange (\autoref{thm:exchange}), and for the reverse direction exchange and \autoref{lm:exchange-exchange} to show that exchanging two elements in $P$ twice leaves $P$ unchanged. \qedhere
  \end{itemize}
\end{proof}

\paragraph*{Substitution}
\autoref{thm:substitution-generalization} proves a generalised version of substitution, where the substitition $\subst{P}{j}{i}$ is on any variable $i$.
\autoref{thm:substitution} instantiates the generalised version to the concrete case where $i$ is the most recently introduced variable $\constr{0}$, as required by subject reduction.

@@

\paragraph*{Subject Reduction}
states that if $P$ is well typed and it reduces to $Q$, then $Q$ is well typed.
In the \picalc{} we distinguish between a reduction $P \reduce{\constr{internal}} Q$ on a channel internal to $P$, and a reduction $P \reduce{\constr{external} \; i} Q$ on a channel $i$ external to $P$ (refer to \autoref{operational-semantics}).
